---
editor: 
  markdown: 
    wrap: 72
---

# Results

## **Passenger Load/Trips Analysis**

```{r}
#| echo: false
#| message: false
#| warning: false

library(lubridate)
library(tidyverse)
library(ggplot2)
library(ggridges)
library(GGally)
library(plotly)
```

```{r}
#| echo: false
#| message: false
#| warning: false

path = "~/MTA_NYCT/MTA_NYCT_Express_Bus_Capacity__Beginning_2023.csv"
bus_load = read.csv(path)
bus_load$Week <- lubridate::mdy(bus_load$Week)
```

### Boxplot: Load Percentage by Direction

```{r}
#| echo: false
#| message: false
#| warning: false

# Compare load percentages by direction
average_load_by_direction <- bus_load %>%
  group_by(Direction) %>%
  summarize(Average_Load = mean(Load.Percentage, na.rm = TRUE))


# Create boxplots of load percentages by direction
ggplot(bus_load, aes(x = Direction, y = Load.Percentage, fill = Direction)) +
  geom_boxplot() + # Use geom_boxplot for boxplots
  labs(
    title = "Average Load Percentage by Direction",
    x = "Direction",
    y = "Load Percentage (%)"
  ) +
  theme_minimal() +
  scale_fill_brewer(palette = "Set1") # Use a color palette for differentiation
```

#### **Comment**:

**Median Load Percentage**: The median load percentage seems relatively
different. NB and SB are the first 2 largest Load Percentage while WB
and EB are 2 smalless, Which indicates North and South may stand more
passengers while the other 2 direction have lower utilization.

**Outliers**: All directions seems a lot of outliers, which are not
taken into calculation consideration, I think we could further explore
those and remove potential outliers which may influence our results.

**Variation**: By seeing the interquantile of those directions, we can
see that SB direction have more variation with respect to the load
percentage, which indicates south direction should be explored further.

### Horizontal barchart: Avg. Load Percentage per Borough

```{r}
#| echo: false
#| message: false
#| warning: false

average_load_by_hour_borough <- bus_load |>
  group_by(Borough) |>
  summarize(Average_Load = mean(Load.Percentage, na.rm = TRUE))


average_load_by_hour_borough |> ggplot(aes(x = fct_reorder(Borough, Average_Load, median), y = Average_Load)) +
  geom_bar(stat = 'identity') + 
  labs(title = 'Average Load Percentage per Borough',
       x = "Borough", 
       y = "Avg. Load") +
  scale_fill_brewer(palette = "Set1")+
  coord_flip()
```

#### **Comment**:

**Average Load**: Staten Island shows the highest average load
percentage. This could indicate a higher utilization of the
transportation system, or possibly fewer transportation options leading
to more crowded buses or trains.

### Horizontal barchart: Avg. Trips with APC per Borough

```{r}
#| echo: false
#| message: false
#| warning: false

average_load_by_hour_borough <- bus_load |>
  group_by(Borough) |>
  summarize(Average_Trip = mean(Trips.with.APC, na.rm = TRUE))


average_load_by_hour_borough |> ggplot(aes(x =  fct_reorder(Borough, Average_Trip, median), y = Average_Trip)) +
  geom_bar(stat = 'identity') + 
  labs(title = 'Average Trips per Borough',
       x = "Borough", 
       y = "Avg. Trip") +
    scale_fill_brewer(palette = "Set1")+
  coord_flip()
```

#### **Comment**:

**Average Trip**: Staten Island stands out with a significantly higher
average number of trips. This could indicate a higher frequency of trips
or that fewer routes are being used more frequently. Therefore, we
should combine with Average load to see it's the bad utilization or we
should actually increase more routes.

### Cleveland plot: Avg. Load Percentage per Route

```{r, fig.width=5, fig.height=11}
#| echo: false
#| message: false
#| warning: false

# Calculate the average load by route and borough
average_load_by_route <- bus_load |>
  group_by(Route, Borough) |>
  summarise(Average_Load = mean(Load.Percentage, na.rm = TRUE), .groups = 'drop')

# Now, create the Cleveland dot plot
average_load_by_route |>
  ggplot(aes(x = Average_Load, y = fct_reorder(Route, Average_Load), color = Borough)) +
  geom_point(size = 2) +  # Adjust the size of the points to make them clearer
  scale_color_brewer(palette = "Set3") +  # Use a color palette that's suitable for discrete variables
  ggtitle("Average Load per Route by Borough") +
  ylab("Route") +
  xlab("Average Load Percentage") +
  theme_linedraw() +
  theme(
    legend.position = "right"  # Adjust legend position
  ) +
  facet_grid(rows = vars(Borough), scales = "free", space = "free") +
  guides(color = guide_legend(title = "Borough"))  # Add title to the legend
```

#### **Comment**:

**Route Load Variation**: There is a wide variation in the average load
percentages across different routes. Some routes have a low average
load, while others are approaching 50%, which indicates differences in
route popularity, capacity, or possibly the time of day the data was
collected.

**Comparison Across Boroughs**: It appears that Staten Island has
several routes with higher load percentages, while routes in Queens have
a lower average load. This could reflect differences in transportation
needs or service levels across boroughs.

## **Time-based Trends**

### Line plot: Avg. Load Percentage of NYC Bus by Hour

```{r}
#| echo: false
#| message: false
#| warning: false

# Calculate average load percentage by hour
average_load_by_hour <- bus_load |>
  group_by(Hour) |>
  summarize(Average_Load = mean(Load.Percentage, na.rm = TRUE))

highest_load <- average_load_by_hour[which.max(average_load_by_hour$Average_Load), ]
lowest_load <- average_load_by_hour[which.min(average_load_by_hour$Average_Load), ]

average_load_by_hour |>
  ggplot(aes(x = Hour, y = Average_Load)) +
  geom_line(color = "blue", size = 1) + 
  geom_point(color = "red", size = 2) + 
  labs(
    title = "Average Load Percentage of NYC Bus (by Hour)",
    x = "Hour of the Day (h)",
    y = "Average Load Percentage (%)"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(size = 12, face = "bold"),
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12),
    axis.text.x = element_text(angle = 90, vjust = 0.5),
    axis.text.y = element_text(color = "blue"),
    legend.position = "none"
  ) +
  scale_x_continuous(breaks = seq(0, 23, by = 1)) +
  scale_y_continuous(labels = scales::percent) +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank()
  ) +
  # Annotate highest load percentage
  geom_text(data = highest_load, aes(label = paste("Highest: ", scales::percent(Average_Load))), 
            nudge_y = 0.02, size = 4, hjust = 0) + 
  # Annotate lowest load percentage
  geom_text(data = lowest_load, aes(label = paste("Lowest: ", scales::percent(Average_Load))), 
            nudge_y = -0.02, size = 4, hjust = 0)
```

#### **Comment**:

**Peak Times**: There is a clear peak in average load percentage around
the 8th hour marked as the highest load at 48%, and another small peak
around 16-17th hour. This likely corresponds to morning rush hour, a
common time for people to commute to work or school, and the night rush
hour, a common time for going home after work.

**Lowest Load**: The graph points out the lowest load percentage at 18%,
which occurs around the 4th hour, likely in the very early morning when
the least number of people are traveling.

**Nighttime Load**: There is an uptick in load percentage later in the
evening, after the 20th hour, which could correspond to people returning
home from late shifts or social activities.

### Barchart: Avg. Trips with APC of NYC Bus by Hour

```{r}
#| echo: false
#| message: false
#| warning: false

# Calculate average trips with APC by hour
average_trips_with_APC_by_hour <- bus_load |>
  group_by(Hour) |>
  summarize(Average_Trips_with_APC = mean(Trips.with.APC, na.rm = TRUE))


# Plotting the data
ggplot(average_trips_with_APC_by_hour, aes(x = Hour, y = Average_Trips_with_APC)) +
  geom_bar(stat = "identity", fill = "steelblue") + # Use geom_bar for bar plots
  labs(
    title = "Average Trips with APC of NYC Bus (by Hour)",
    x = "Hour of the Day (h)",
    y = "Average # of Trips"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 12, face = "bold"),
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12),
    axis.text.x = element_text(angle = 90, vjust = 0.5) # Rotate the x-axis labels for better readability
  ) +
  scale_x_continuous(breaks = seq(0, 23, by = 1)) # Assuming 'Hour' ranges from 0 to 23
```

#### **Comment**:

**Peak Hours**: The chart shows distinct peaks in the number of trips
during certain hours of the day. The most prominent peak appears to be
in the morning hours, around 8 AM, suggesting a morning rush hour with
the highest average number of trips. There's another less pronounced
peak in the late afternoon or early evening, possibly indicating an
evening rush hour.

**Off-Peak Variation**: Outside of the peak hours, the number of trips
decreases but fluctuates throughout the day. Midday hours show a
moderate level of trips, which could be due to people traveling for
non-work related reasons or staggered work schedules.

**Nighttime Service**: There's a noticeable drop in the average number
of trips late at night, reaching the lowest point in the early morning
hours around 4 AM to 5 AM, which is expected as this is typically
outside of standard commuting times.

### Ridgeline plot: Load Percentage Distribution by Day Type

```{r}
#| echo: false
#| message: false
#| warning: false

bus_load |>
  ggplot(aes(x = Load.Percentage, y = Day.Type)) +
  geom_density_ridges(fill = 'blue') +  # Add gradient and adjust the scale and minimum height
  labs(
    title = "Load Percentage Distribution by Day Type",
    x = "Percentage of Load (%)",
    y = "Day Type"
  ) +
  theme_ridges(font_size = 12, grid = TRUE)  # Use the ridges theme with adjusted font size and grid lines
```

#### **Comment**:

**Distribution Shape**: The distribution of load percentages on weekdays
has a wider spread than on weekends, suggesting there is more variation
in bus load percentages during weekdays.

**Peaks and Load**: There appear to be multiple peaks for both weekdays
and weekends. This could indicate that there are specific times during
the day when buses are consistently more or less loaded. The highest
peak on weekdays suggests a time when buses are most consistently at a
similar load percentage, possibly during rush hours.

**Weekday vs. Weekend Trends**: The weekday distribution has a higher
peak, which might indicate a more consistent pattern of high load
percentages, likely associated with commuter traffic. The weekend
distribution is broader with a less pronounced peak, suggesting a more
varied load percentage, which could be due to a more irregular or
leisure-based travel pattern.

**Bus Utilization**: The graph could suggest that buses are more heavily
utilized during weekdays as compared to weekends, which would align with
typical workweek patterns.

### Ridgeline plot: Trip with APC Distribution by Day Type

```{r}
#| echo: false
#| message: false
#| warning: false

bus_load |>
  ggplot(aes(x = log(Trips.with.APC, 2), y = Day.Type)) +
  geom_density_ridges(fill = 'blue') +  # Add gradient and adjust the scale and minimum height
  labs(
    title = "Log2(Trips with APC Distribution) by Day Type",
    x = "Log2(Trips with APC)",
    y = "Day Type"
  ) +
  theme_ridges(font_size = 12, grid = TRUE)  # Use the ridges theme with adjusted font size and grid lines
```

#### **Comment**:

**Weekend Peaks**: The distribution for weekends has pronounced peaks,
especially the very sharp peak at the lower end of the x-axis,
indicating that there are many instances of a specific, lower number of
trips (possibly single or very few trips on certain routes). This
suggests that there's less variability in the number of trips on
weekends, with certain trip counts being far more common than others.

**Weekday Distribution**: The weekday distribution is smoother and more
spread out, which indicates a greater variability in trip counts. This
could be due to a more uniform distribution of commuter travel spread
throughout the workday.

**Use of Logarithmic Scale**: By using a logarithmic scale (log2), the
graph allows for easier visualization and comparison of data across a
wide range of values. It's especially useful when dealing with
exponential growth or large disparities in data.

**Low Trip Counts**: The first peak in the weekend distribution suggests
that the lowest trip counts are more common on weekends than on
weekdays. This could be due to reduced service schedules or lower
demand.

## **Multiple variable Interactions**

### Barcharts by Borough: Avg. Load Percentage with respect to Hour(Each Borough)

```{r}
#| echo: false
#| message: false
#| warning: false

average_load_by_hour_borough <- bus_load |>
  group_by(Borough, Hour) |>
  summarize(Average_Load = mean(Load.Percentage, na.rm = TRUE))

average_load_by_hour_borough |>
  ggplot(aes(x = Hour, y = Average_Load)) +
  geom_bar(stat = 'identity', fill = 'blue') + 
  facet_wrap(~Borough) + 
  labs(
    title = "Average Load Percentage with respect to each Borough",
    x = "Hour of the Day (h)",
    y = "Average Load Percentage (%)"
  )
```

#### **Comment**:

**Variability**: Each borough exhibits variability in load percentages
throughout the day. This could be reflective of different travel
patterns, with peaks likely corresponding to rush hours.

**Peak Times**: Bronx: Shows pronounced peaks, which could suggest rush
hours are more pronounced in the Bronx, with higher bus usage during
these times.

**Brooklyn**: Exhibits a more even distribution with less pronounced
peaks, possibly indicating a more steady usage of buses throughout the
day.

**Queens**: Similar to the Bronx, there are distinct peaks that may
correspond to rush hours.

**Staten Island**: Has a more varied pattern with multiple peaks,
suggesting several busy periods throughout the day.

### Scatterplot: Correlation between Load Percentage and Trips

```{r}
#| echo: false
#| message: false
#| warning: false

sample <- slice_sample(bus_load ,prop=0.1) #sample data


p <- ggplot(sample, aes(x = Load.Percentage, y = Trips.with.APC, text = Borough)) + 
  geom_point(alpha = 0.4) + 
  labs(title = "Correlation between Load Percentage and Trips ScatterPlot",
       x = "Load percentage (%)", 
       y = "Trips with APC")
ggplotly(p, tooltip = "text")
```

#### **Comment**:

**Data Distribution**: A scatter plot with a lot of data points
clustered at the lower end of the load percentage axis might suggest
that most bus trips have a low to moderate load percentage. But the
correlation relationship is not so apparent.

**Outliers**: There are points scattered more widely along the y-axis at
higher load percentages (representing the Staten Island Borough), these
could be considered outliers indicating occasional trips with
significantly higher load percentages.

### Stacked bar chart: Direction composition of each borough

```{r}
#| echo: false
#| message: false
#| warning: false

bus_load |> 
  ggplot(aes(x = Borough, fill = Direction)) + 
  geom_bar(position = "stack", width = 0.7) +  # Adjust bar width here if necessary
  scale_fill_brewer(palette = "Set1") +  # This will use a set of distinct colors
  labs(
    title = "Distribution of Directions by Borough",
    x = "Borough",
    y = "Count",
    fill = "Direction"  # Label for the legend
  ) +
  theme_minimal() +  # Using a minimal theme for clarity
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),  # Adjust text angle and horizontal justification
    legend.position = "bottom"  # Position the legend at the bottom
  )
```

#### **Comment**:

**Graph distribution**: From the graph, we know that only Brooklyn has 4
directions of routes, while Bronx and Staten Island have only North and
South direction. Queens has only East and West. According to previous
graphs, it indicates that North and South have higher load percentages.
This is fit for that Bronx and Staten Island have higher load percentage
while Queens has lower percentage.

## **Non-technical summary**:

**Busyness by Time and Place**: Some graphs showed how busy the buses
are at different times of the day in various boroughs. For example,
there are times in the morning and evening when the buses are
particularly full, likely when people are going to or from work.

**Direction Matters**: Other charts indicated that in some boroughs,
buses going in certain directions are busier than others. This might
tell us where more people work or live, or maybe where popular
destinations are.

**Weekdays vs. Weekends**: There's also a difference between weekdays
and weekends. Buses are consistently busier on weekdays, probably due to
people's work schedules, while weekends show more varied bus usage,
reflecting more leisure or irregular travel.

**Different Boroughs, Different Patterns**: We know that Staten Island
is the Borough with highest load percentage and Trips with APC. While
Bronx also have higher Load percentage, it have fewer Trips with APC,
which indicates it may not utilize the transportation fully. Queens have
the least load percentage, which means we may need to give less
resources to Queens, and add resources to Staten Island.

**Relationship**: Because there's not clear relationship between Load
percentage and Trips with APC. Therefore, we should combine both to
answer the question of Route optimization and further policy and
planning.


## References:
  
**1** https://stackoverflow.com/questions/41940439/display-multiple-d3-js-charts-in-a-single-html-page

**2** https://datawanderings.com/2019/11/01/tutorial-making-an-interactive-line-chart-in-d3-js-v-5/

**3** https://d3-graph-gallery.com/graph/interactivity_button.html

**4** https://forum.freecodecamp.org/t/d3-tooltip-cant-access-data-shows-undefined/440547


